\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-21 papers/jfp/4-schedulers.tex
Proofreading

As per the definition of minimality~\ref{def-minimal}, a minimal build
system must \textbf{rebuild only out-of-date keys} and at most once. The only
way to achieve the ``at most once'' requirement while producing a correct build
result (\S\ref{sec-build-correctness}) is to \textbf{build all keys in an
order that respects their dependencies}.

@@

The topological scheduler pre-computes a linear order of tasks, which when
followed ensures dependencies are satisfied, then executes the required tasks in
that order. Computing such a linear order is straightforward -- given a task
description and a target \hs{key}, first find the (acyclic) graph of the
\hs{key}'s dependencies, then compute a topological order. Taking the \Make
example from Fig.~\ref{fig-make}, we might compute the order:

@@

To handle dynamic dependencies we cannot precompute a static order -- we must
interleave running tasks and ordering tasks. One approach is just to build tasks
in an arbitrary order, and if a task calls \hs{fetch} on an out-of-date key
\cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
from Fig.~\ref{fig-make}, we might build the tasks in the order:

@@

\noindent
We start with \cmd{main.exe} (chosen randomly), but discover it depends on
\cmd{main.o}, so instead start building \cmd{main.o}. Next we choose to build
\cmd{util.o} (again, randomly), before finally returning to \cmd{main.exe} that
now has all its dependencies available and completes successfully.

The approach works, but has a number of disadvantages. Firstly, it requires a
technical mechanism to abort a task, which is easy in our theoretical setting
with \hs{Task} (see an implementation in~\S\ref{sec-implementation-excel}) but
leads to engineering concerns in the real world. Secondly, it is not minimal in
the sense that a task may start, do some meaningful work, and then abort,
repeating that same work when restarted.

>>>>

As per the definition of minimality~(\S\ref{def-minimal}), a minimal build
system must \textbf{rebuild only out-of-date keys} and at most once. The only
way to achieve the ``at most once'' requirement while producing a correct build
result (\S\ref{sec-build-correctness}) is to \textbf{build all keys in an
order that respects their dependencies}.

@@

The topological scheduler pre-computes a linear order of tasks, which when
followed ensures dependencies are satisfied, then executes the required tasks in
that order. Computing such a linear order is straightforward -- given a task
description and a target \hs{key}, first find the (acyclic) graph of the
\hs{key}'s dependencies, then compute a topological order. Taking the \Make
example from Fig.~\ref{fig-make}, we might compute the following order:

@@

To handle dynamic dependencies we cannot precompute a static order -- we must
interleave running tasks and ordering tasks. One approach is just to build tasks
in an arbitrary order, and if a task calls \hs{fetch} on an out-of-date key
\cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
from Fig.~\ref{fig-make}, we might build the tasks as follows:

@@

\noindent
We start with \cmd{main.exe} (an arbitrary choice), but discover it depends on
\cmd{main.o}, so instead start building \cmd{main.o}. Next we choose to build
\cmd{util.o} (again, arbitrarily), before finally returning to \cmd{main.exe}
that now has all its dependencies available and completes successfully.

This approach works, but has a number of disadvantages. Firstly, it requires a
technical mechanism to abort a task, which is easy in our theoretical setting
with \hs{Task} (see an implementation in~\S\ref{sec-implementation-excel}) but
leads to engineering concerns in the real world. Secondly, it is not minimal in
the sense that a task may start, do some meaningful work, and then abort,
repeating that same work when restarted.

### 2019-10-21 papers/jfp/6-implementations.tex
Proofreading

We can define schedulers and rebuilders more concretely with the types
(Fig.~\ref{fig-types}):

@@

These two abstractions are the key to modularity: \emph{we can combine any
scheduler with any rebuilder, and obtain a correct build system}.
In this section we will write a scheduler for each column of
Table~\ref{tab-build-systems}, and a rebuilder for each row; then combine them
to obtain the build systems in the table's body.

@@

As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step
traces, but here we choose to focus on the more explicit verifying traces. We
have implemented verifying step traces in our framework, and they compose with
schedulers as you would hope.

@@

The \Bazel build system uses a restarting scheduler whose implementation we
omit. It is similar to \Excel's \hs{restarting} scheduler defined in
Fig.~\ref{fig-excel-implementation}, but instead of building keys in the order
specified by the persistently stored calc chain, \Bazel uses a \emph{build
queue}. The build starts with the queue containing all dirty keys. Similar to
\Excel, the rebuilding of a key extracted from the queue may fail because one of
its dynamic dependencies is dirty. In this case the key is marked as
\emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
rebuilt, all keys that were previously blocked on it are added back to the
queue, and their build is eventually restarted.

@@

Using the abstractions built thus far, we have shown how to combine schedulers
with rebuilders to reproduce existing build systems. To us, the most interesting
build system as yet unavailable would combine a suspending scheduler with
constructive traces~--~providing a cloud-capable build system that is minimal,
and supports both early cutoff and monadic dependencies. Using our framework it
is possible to define and test such a system, which we call \Cloud \Shake. All
we need to do is combine \hs{suspending} with \hs{ctRebuilder},
as shown in Fig.~\ref{fig-cloud-implementations}.

>>>>

We can define schedulers and rebuilders more concretely with the following types
(Fig.~\ref{fig-types}):

@@

These two abstractions are the key to modularity: \emph{we can compose any
scheduler with any rebuilder, and obtain a correct build system}.
In this section we will write a scheduler for each column of
Table~\ref{tab-build-systems}, and a rebuilder for each row; then compose them
to obtain the build systems in the table's body.

@@

As discussed in \S\ref{sec-step-traces}, \Shake actually uses verifying step
traces, but here we choose to focus on the more explicit verifying traces. We
have implemented verifying step traces in our framework, and they compose with
schedulers as you would hope.

@@

The \Bazel build system uses a restarting scheduler whose implementation we
omit. It is similar to \Excel's \hs{restarting} scheduler defined in
Fig.~\ref{fig-excel-implementation}, but instead of building keys in the order
specified by the persistently stored calc chain, \Bazel uses a \emph{build
queue}. The build starts with the queue containing all dirty keys. Similarly to
\Excel, the rebuilding of a key extracted from the queue may fail because one of
its dynamic dependencies is dirty. In this case the key is marked as
\emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
rebuilt, all keys that were previously blocked on it are added back to the
queue, and their build is eventually restarted.

@@

Using the abstractions built thus far, we have shown how to compose schedulers
with rebuilders to reproduce existing build systems. To us, the most interesting
build system as yet unavailable would compose a suspending scheduler with
constructive traces~--~providing a cloud-capable build system that is minimal,
and supports both early cutoff and monadic dependencies. Using our framework it
is possible to define and test such a system, which we call \Cloud \Shake. All
we need to do is compose \hs{suspending} with \hs{ctRebuilder},
as shown in Fig.~\ref{fig-cloud-implementations}.

### 2019-10-21 papers/jfp/7-experience.tex
Proofreading

The effect of using a concrete design language went well beyond
merely \emph{expressing} our ideas: it \emph{directly influenced}
our thinking.  For example, here are definitions for a scheduler and rebuilder,
from \S~\ref{sec-concrete-impl}:

@@

The original design of \Shake has not changed since the initial paper, but the
implementation has continued to mature -- there have been roughly 5,000
subsequent commits to the \Shake
project\footnote{See \url{https://github.com/ndmitchell/shake}.}. These commits
add concepts like \emph{resources} (for handling situations when two build tasks
contend on a single external resource), rewriting serialisation to be faster,
documentation including a website\footnote{See \url{https://shakebuild.com}.},
and add a lot of tests. The biggest change in that time period was an
implementation change: moving from
blocking threads to continuations for the suspending scheduler.
But otherwise, almost all external and internal details remain the
  same\footnote{The most visible change is purely notational: switching from
  \hs{*>} to \hs{@\%@>} for defining rules, because a conflicting \hs{*>}
  operator was added to the Haskell \hs{Prelude}.}.
We consider the lack of change suggestive that \Shake is based on fundamental
principles -- principles we can now name and describe as a consequence
of this paper.

There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
in the original paper the tasks (there called ``actions'')
were described in continuation-passing style using the data
type\footnote{The original paper uses concrete types \hs{Key} and \hs{Value}.
Here we generalise these types to~\hs{k} and~\hs{v}, and also add~\hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}.}:
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{0.5mm}

@@

\subsection{Experience from using \Cloud \Shake}\label{sec-using-cloud-shake}

While we expected the GHC build system to be the first to take advantage of
\Cloud \Shake, we were actually beaten to it by Standard Chartered who
report\footnote{\mbox{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}}:

@@

\begin{itemize}
\item \textbf{Irrelevant differences}: A common problem is that you do not get
      shared caching when you want it. As one example, imagine two users install
      \cmd{gcc} on different paths (say \cmd{/usr/bin/gcc} and \cmd{/usr/local/bin/gcc}).
      If these paths are recorded by the build system, the users won't share
      cache entries. As another example, consider a compiler that embeds the current time
      in the output: any users who build that file locally won't get any shared
      caching of subsequent outputs. Possible solutions include using relative paths;
      depending only on version numbers for system binaries (e.g. \cmd{gcc});
      controlling the environment closely (e.g. using \Nix); and extra flags to
      encourage compilers to be more deterministic.
\item \textbf{Insufficient produced files}: A build rule must declare all files
      it produces, so these can be included in the cache.
      As an example using Haskell, compilation of \cmd{Foo.hs} produces
      \cmd{Foo.hi} and \cmd{Foo.o}. If you declare the rule as producing \cmd{Foo.hi},
      and other rules \emph{depend on} \cmd{Foo.hi}, but \emph{also use} \cmd{Foo.o} after
      depending on \cmd{Foo.hi}, a local build will probably work (although treating \cmd{Foo.o}
      as a proper dependency would definitely be preferable). However,
      if \cmd{Foo.hi} is downloaded from a remote cache, \cmd{Foo.o} will not be present,
      and subsequent commands may fail (e.g. linking).
      In practice, most issues encountered
      during the move to cloud builds for GHC were caused by failing to declare produced
      files.
\item \textbf{Missing dependencies}: While missing dependencies are
      always a problem, the move to a cloud build makes them more serious.
      With local builds outputs will be built at least once per user, but with a
      cloud build they might only be built \emph{once ever}.
\end{itemize}

To help with the final two issues -- insufficient dependencies and produced
files -- we have further enhanced the \Shake lint modes, coupling them to a
program called \cmd{FSATrace}, which detects which files are read/written by a
command line execution. Such information has been very helpful in making the GHC
build cloud ready~\cite{cloud_hadrian}.

\subsection{Experience from building GHC with \Shake}\label{sec-hadrian}

@@

The subsequent development of \Cloud \Shake allows GHC developers to take advantage of
sharing build results between builds, which is particularly promising in the
context of GHC's continuous integration (CI) infrastructure responsible for
building and testing every GHC commit. Building GHC~8.8 from scratch with
default settings takes around 1 hour on Windows using either \Hadrian or the
original \Make-based build system. This time includes building the compiler itself,
29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
only a small part of the whole codebase, making the cloud-build functionality
highly desirable. For example, if a commit does not touch any GHC source files
(e.g. only modifies the documentation), \Hadrian can build GHC from scratch in under
3~minutes, by simply creating symbolic links to the previously build results
stored in the \Cloud~\Shake cache.

>>>>

The effect of using a concrete design language went well beyond
merely \emph{expressing} our ideas: it \emph{directly influenced}
our thinking.  For example, here are definitions for a scheduler and rebuilder,
from~\S\ref{sec-concrete-impl}:

@@

The original design of \Shake has not changed since the initial paper, but the
implementation has continued to mature -- there have been roughly 5,000
subsequent commits to the \Shake
project\footnote{See \url{https://github.com/ndmitchell/shake}.}. These commits
added concepts like \emph{resources} (for handling situations when two build tasks
contend on a single external resource), rewriting serialisation to be faster,
documentation including a website\footnote{See \url{https://shakebuild.com}.},
and add a lot of tests. The biggest change in that time period was an
implementation change: moving from
blocking threads to continuations for the suspending scheduler.
But otherwise, almost all external and internal details remain the
  same\footnote{The most visible change is purely notational: switching from
  \hs{*>} to \hs{@\%@>} for defining rules, because a conflicting \hs{*>}
  operator was added to the Haskell \hs{Prelude}.}.
We consider the lack of change suggestive that \Shake is based on fundamental
principles -- principles we can now name and describe as a consequence
of this paper.

There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
in the original paper the tasks (there called ``actions'')
were described in continuation-passing style using the following data
type\footnote{The original paper uses concrete types \hs{Key} and \hs{Value}.
Here we generalise these types to~\hs{k} and~\hs{v}, and also add~\hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}.}:
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{0.5mm}

@@

\subsection{Experience from Using \Cloud \Shake}\label{sec-using-cloud-shake}

While we expected the GHC build system to be the first to take advantage of
\Cloud \Shake, we were actually beaten to it by Standard Chartered who
reported\footnote{\mbox{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}}:

@@

\begin{itemize}
\item \textbf{Irrelevant differences}: A common problem is that you do not get
      shared caching when you want it. As one example, imagine two users install
      \cmd{gcc} on different paths (say \cmd{/usr/bin/gcc} and \cmd{/usr/local/bin/gcc}).
      If these paths are recorded by the build system, the users won't share
      cache entries. As another example, consider a compiler that embeds the current time
      in the output: any users who build that file locally won't get any shared
      caching of subsequent outputs. Possible solutions include using relative paths;
      depending only on version numbers for system binaries (e.g. \cmd{gcc});
      controlling the environment closely (e.g. using \Nix); and extra flags to
      encourage compilers to be more deterministic.
\item \textbf{Insufficient produced files}: A build rule must declare all files
      it produces, so these can be included in the cache.
      As an example using Haskell, compilation of \cmd{Foo.hs} produces
      \cmd{Foo.hi} and \cmd{Foo.o}. If you declare the rule as producing \cmd{Foo.hi},
      and other rules \emph{depend on} \cmd{Foo.hi}, but \emph{also use} \cmd{Foo.o} after
      depending on \cmd{Foo.hi}, a local build will probably work (although treating \cmd{Foo.o}
      as a proper dependency would definitely be preferable). However,
      if \cmd{Foo.hi} is downloaded from a remote cache, \cmd{Foo.o} will not be present,
      and subsequent commands may fail (e.g. linking).
      In practice, most issues encountered
      during the move to cloud builds for GHC were caused by failing to declare produced
      files.
\item \textbf{Missing dependencies}: While missing dependencies are
      always a problem, the move to a cloud build makes them more serious.
      With local builds, outputs will be built at least once per user, but with a
      cloud build they might only be built \emph{once ever}.
\end{itemize}

To help with the final two issues -- insufficient dependencies and produced
files -- we have further enhanced the \Shake lint modes, coupling them to a
utility called \cmd{FSATrace}, which detects which files are read/written by a
command line execution. Such information has been very helpful in making the GHC
build cloud ready~\cite{cloud_hadrian}.

\subsection{Experience from Building GHC with \Shake}\label{sec-hadrian}

@@

The subsequent development of \Cloud \Shake allows GHC developers to take advantage of
sharing build results between builds, which is particularly promising in the
context of GHC's continuous integration (CI) infrastructure responsible for
building and testing every GHC commit. Building GHC~8.8 from scratch with
default settings takes around 1 hour on Windows using either \Hadrian or the
original \Make-based build system. This time includes building the compiler itself,
29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
only a small part of the whole codebase, making the cloud-build functionality
highly desirable. For example, if a commit does not touch any GHC source files
(e.g. only modifies the documentation), \Hadrian can build GHC from scratch in under
3~minutes, by simply creating symbolic links to the previously built results
stored in the \Cloud~\Shake cache.

### 2019-10-21 papers/jfp/8-engineering.tex
Proofreading

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition~(\S\ref{sec-build-correctness}): instead of requiring that the
produced value \emph{equals the result} of recomputing the task, we now require
that produced value \emph{belongs to the set of possible results} of
recomputing the task, i.e. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
example.

@@

Note that like all key-dependent tasks, this example could be expressed
without key-dependence, at the cost of some type safety. As we'll see in
\S\ref{sec-multiple-outputs} and \S\ref{sec-tracking-aspects}, using key-dependent
value types can make it easier to work with more complicated tasks.

@@

\subsection{File watching vs polling}\label{sec-file-watching}

>>>>

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition~(\S\ref{sec-build-correctness}): instead of requiring that the
produced value \emph{equals the result} of recomputing the task, we now require
that produced value \emph{belongs to the set of possible results} of
recomputing the task, e.g. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
example.

@@

Note that like all key-dependent tasks, this example could be expressed
without key-dependence, at the cost of some type safety. As we will see in
\S\ref{sec-multiple-outputs} and \S\ref{sec-tracking-aspects}, using key-dependent
value types can make it easier to work with more complicated tasks.

@@

\subsection{File Watching vs Polling}\label{sec-file-watching}

### 2019-10-21 papers/jfp/9-related.tex
Proofreading

\item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
matches \Shake at the level of detail given here, differing only on aspects like
rules producing multiple files~\S\ref{sec-polymorphism}. While \Redo predates
\Shake, they were developed independently; we use \Shake as a prototypical
example of a monadic build system because its implementation presents a closer
mapping to our \hs{Task} abstraction.

@@

\noindent
To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed. The main difference from our
\hs{Tasks} abstraction is that instead of supplying a mapping from keys to
tasks, a \Fabricate script supplies a list of build statements, in a
user-scheduled order, without declaring what each statement reads or write.

>>>>

\item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
matches \Shake at the level of detail given here, differing only in aspects like
rules producing multiple files~\S\ref{sec-polymorphism}. While \Redo predates
\Shake, they were developed independently; we use \Shake as a prototypical
example of a monadic build system because its implementation presents a closer
mapping to our \hs{Task} abstraction.

@@

\noindent
To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed. The main difference from our
\hs{Tasks} abstraction is that instead of supplying a mapping from keys to
tasks, a \Fabricate script supplies a list of build statements, in a
\emph{user-scheduled order}, without declaring what each statement reads or write.
